### 2015-01-22 2014-gimme.tex
renamed to 2014-gimme

Until recently, studies of alternative splicing have been limited to a
small number of genes and isoforms due to the high cost and low throughput
of sequencing expressed sequence tags (ESTs) and full-length cDNA
libraries.  RNA sequencing (RNA-Seq) using deep short-read sequencing
has been used successfully in many studies to gain unprecedented
insight into the complexity of transcriptomes~\cite{nagalakshmi2008transcriptional,mortazavi2008mapping,pan2008deep,Wang:2008ea}.
It has been estimated that, in human, $92-94$\% of multiexon genes
undergo alternative splicing and different isoforms are expressed in
different tissues\cite{Wang:2008ea}.  This suggests that even in human
a large number of splice variants have not been explored.

@@

Despite the small size of sequencing reads, several studies have
detected novel splice junctions based on alignment of reads spanning
putative exon junctions.  To map reads across exon junctions,
reads are split into two parts and each part is mapped to the genome
independently.  A splice junction is then identified based on
alignments of each half of a read that falls between two exons at
exon-intron boundaries.  With this approach, Wang \emph{et al} have
identified a large number of splice junctions that are not annotated
from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.  These novel
splice sites include both canonical and non-canonical splice sites.
Approximately $46\%-75\%$ of canonical splice sites are supported by
ESTs.  Novel splice junctions have different levels of read coverage
suggesting that both high- and low-expressed isoforms are unannotated.
Using a similar approach, Pickrell {\em et al}~\cite{Pickrell:2010gt}
identified more than 150,000 novel canonical splice junctions in lymphoblastoid
cells.  The study also shows that the number of unannotated splice junctions
varies among cells from different human tissues, which suggests tissue-specific
expression of isoforms\cite{Pickrell:2010gt}.
% A majority of unannotated isoforms found in both studies are from alternative
% splicing of annotated exons. Only a small fraction of isoforms contain
% previously unknown exons.
% @CTB is this last point important?

@@

and Table~\ref{unique_sequences_matched_mouse}); this is probably due to the
incomplete nature of the current chicken genome assembly, which is lacking
approximately 5\% of its true gene content~\cite{hillier2004sequence}.
Interestingly, over a hundred regions were present {\em only} in the local
assemblies, suggesting that the local assemblies might be recovering additional
exons.  Significant numbers of unique regions from both global and local
assemblies showed homology to the mouse genome, indicating that at least some
of these unique sequences represented real sequence content.
Figure~\ref{global_vs_local} shows an example of different isoforms detected by
the two assembly methods.

@@

More importantly, we also mapped paired-end reads from technical replicates to
the same gene models, and found that more than 74\% of the paired-ends mapped
concordantly the gene models.  Most of the reads that did not map were either
highly erroneous or contained low-complexity artifactual sequence that probably
originated from sample processing and reverse
transcription~\cite{martin2011next}.  Thus the merged gene models produced by
Gimme represent the significant majority of the assemblable data.

@@

Of 105,461 splice junctions from the gene models, 2,057 (2\%) junctions have no
spliced reads and only 4,626 (4.4\%) junctions have fewer than 4 spliced reads
(Figure~\ref{cdf_splice}).  Note that 710 junctions are in chrUn\_random contigs,
which may have a great number of genome missassemblies. The number of junctions
outside random chromosomes with no reads is 1,347/99,986 (1.3\%).
% @CTB how do you get junctions with no spliced reads!?
% @LP our models have many 2,400 more junctions in random chromosomes than
% cufflinks and 710 of no reads junctions are from random chromosomes. It is
% likely that those junctions are from missassemblies.
% @LP Another reason is it has something to do with variation in read
% mapping between Tophat and bowtie.  Tophat maps spliced reads with less
% stringency because it has to allow gap alignment and it uses -v mode to map
% spliced reads. On the other hand, the default mode for bowtie is -n mode,
% which is the one I used.
% @LP I used a parameter that set max alignments to 100, only one alignment will
% be reported if there are more than 100 alignments. Maybe I should let
% bowtie report all alignments instead.
More than 95\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
real splice junctions.
% @CTB I'm slightly bugged by this statement.  Are we right?!
% @LP the main reason I use four reads as a cutoff because there are four
% samples and if at least one read from each sample map to the junction, it
% will likely be real.  Apparently, it is not always the case, but that is just
% my simple logic.

@@

Of the 105,461 splice junctions in our gene models, 83,560 (79.2\%) are
supported by ESTs or mRNAs from Genbank (Figure~\ref{chick_est_venn}).  This is
especially surprising since our mRNAseq data is from spleen, and most of the
publicly available ESTs or mRNAs are from other tissues. 16,909 splice junctions
are found by either Cufflinks or our gene models, but not both, suggesting a variation in
sensitivity between two methods. Note that this cross-validation suggests that
the 12,188 novel splice junctions {\em not} seen in publicly available ESTs and
mRNAs are also likely to be real splice junctions from spleen.

@@

We next compared the Gimme gene models to those produced by Cufflinks,
another reference-based approach to building gene models from mRNAseq
data\cite{Trapnell:2010kd}.  We also compared the results from both
methods to the ENSEMBL gene annotations, which are produced by a
pipeline that incorporates de novo gene prediction and homology-based
approaches as well as expression data~\cite{flicek2013ensembl}.

@@

When we apply Gimme and Cufflinks to a publicly available mouse mRNAseq data
set, Gimme and Cufflinks recover approximately the same number of splice
junctions already known from ENSEMBL (Figure~\ref{mus_venn}).  However, Gimme
recovers a substantial number of additional splice variants beyond Cufflinks and
ENSEMBL both.

@@

As shown above, Cufflinks and the Gimme method detect a number of distinct but
equally valid splice junctions, which suggests that we could obtain greater
sensitivity to exon-exon junctions in our gene models by merging both sets of
predictions.  We therefore used Gimme to incorporate the Cufflinks models to
global and local assembly, Table.~\ref{genes_transcripts}.  This resulted in a
decreased number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Figure~\ref{}). The merged gene models recover 49.3\% and 56.2\% of
splice junctions from ESTs and ENSEMBL respectively
(Figure~\ref{combined_venn}), which is about 10\% greater than that from
corresponding unmerged gene models.

@@

To validate our predicted isoforms, we extracted putative coding
sequences from our gene models with ESTScan\cite{Iseli:1999vd}.
ESTScan successfully translated 28,772 of 34,800 (82.6\%) of
isoforms from Gimme models and 28,918 of 38,307 (75.5\%) of
isoforms from Cufflinks to protein sequences with 50 or more
amino acids.  We then searched for homologous sequences in mouse
ENSEMBL, and found that 22,991 (79.9\%) and 22,460 (77.6\%) of
isoforms from Gimme and Cufflinks match mouse proteins at a
bit-score/length $\ge1.0$ (Figure~\ref{bitscore}A).  The number of
distinct genes are 12,945 and 16,731 from Gimme and Cufflinks
respectively.  A bit-score/length ratio greater than 1.0 indicates
that approximately more than half of the protein sequence matches
mouse proteins well.

\subsubsection*{Merged gene models have more complete
transcripts}

The size of genes and isoforms from Gimme are slightly greater
than those from Cufflinks (Figure~\ref{bitscore}B) while the
bit-score/length ratios are slightly lower. To validate that
Gimme software can be used to combine different sets of gene
models to create an improved set of gene models, we used our
software to combine Gimme and Cufflinks models to create merged
gene models and found that the size of transcripts from merged
gene models are greater than those from either Cufflinks or Gimme
models.  Total of 38,535 of 45,793 (84.2\%) isoforms were
successfully translated to protein sequences with 50 or more
amino acids and 31,765 (82.4\%) from 13,958 genes matched mouse
proteins. A majority of genes and isoforms also have a
bit-score/length ratio greater than 1.0.  This suggests that
merging Cufflinks and Gimme models creates more complete gene
models.

% @CTB for discussion: what's surprising about all of this? so many splice
% variants, etc?  Note we need to go over pickrell paper carefully.  @LP I
% discuss about it below. Hope it helps.

@@

\subsection*{{\em De novo} assembly should be used to extend gene models from
Cufflinks}
For organisms with a reference genome and ENSEMBL annotation, Cufflinks with
ENSEMBL gene models as a reference guide seems to be a preferable method for
building gene models. However, we have shown that ENSEMBL models in chicken are
missing a substantial number of splice variants found in ESTs and mRNAs that
are also not recovered by Cufflinks. Moreover, results from mouse data suggest
that Tophat's default parameters may have been tailored to maximize the finding
of isoforms in ENSEMBL models (Figure~\ref{mus_venn}), which could lead to
restricted detection of novel splice variants.
% 85,512 (97.2\%) of splice
% junctions found by Cufflinks are also canonical compared to 92.6\% in mouse
% ENSEMBL gene models. Assuming that mouse data were not selected for splice
% variants with canonical splice junctions, the percentage of canonical splice
% sites found by Cufflinks is rather high. 
Therefore, gene models built from RNA-Seq reads using Cufflinks + ENSEMBL gene
models alone may not be suitable for genes and isoforms analyses that include
finding novel genes and splice variants.  {\em De novo} assembly is not tied to
any specific gene model parameters and does not only detect canonical splice
sites, thus we recommend that it is used to extend gene models built from
Cufflinks + ENSEMBL gene models, especially in organisms with relatively
incomplete ENSEMBL gene models.  Besides {\em de novo} assembly,
annotation from other sources can also be incorporated to extend Cufflinks models
using our pipeline.

@@

\subsection*{Mapping Reads to The Genome and Gene Models}

@@

% @LP check the version of tophat
Reads from each dataset were first assembled separately by global assembly
without using a reference genome.  In contrast, reads from each dataset were
first mapped to the chicken genome using Tophat2.  Then only reads mapped to
the genome were assembled by chromosome in the local assembly
(Figure~\ref{local_assembly}).  Global and local assembly was performed using
Velvet version 1.2.03\cite{Zerbino:2008vu} with default parameters except for
hash length (k-mer).  A range of k-mer length from 21-31 was used to assemble
reads from chicken data and k-mer length 27 was used to assemble reads from
mouse data.  Lastly, transcripts from both methods were assembled by Oases
version 0.2.06\cite{Schulz:2012je}.

@@

Figure~\ref{img:pipeline} depicts an overall gene model construction
pipeline.  Transcripts of all datasets from local and global assembly were
mapped to the chicken genome using BLAT\cite{Kent:2002tv}. Alignments and gaps
from BLAT outputs are considered exons and introns respectively.  Optionally,
data from other sources (ESTs, RefGenes, etc.) can be incorporated with
transcripts from the assembly to improve gene models.  All transcripts are then
assembled using Gimme, a program that assembles transcripts based on their
alignments to the reference genome.  An algorithm for assembling transcripts is
described below.  A maximum set of transcripts obtained from Gimme are then
reduced to only a minimum set of transcripts that contain all splice junctions
and untranslated regions (UTRs).

@@

The algorithm first builds an intron graph using introns as nodes.  Each intron
contains exons whose splice sites perfectly match intron boundaries. At the
$5\prime$ and the $3\prime$ ends, some exons may be smaller than others due to
incomplete assembly (supp.Fig.??). Only the largest exons are kept to ensure
that the transcript has the most complete exons (Figure~\ref{algorithm}).
Transcripts were then grouped into the same gene if they have at least one
intron or exon in common.  Then, a splice graph composed of exons is created
and structures of isoforms are derived from traversing paths in the splice
graph.  Gimme is open-source and available at
\texttt{https://github.com/ged-lab/gimme}.

@@

\section*{Figure Legends}

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{mapped-reads.pdf}
\end{center}
\caption{
{\bf Percentages of reads mapped to the gene models from Ensembl
and Gimme.}
Reads were mapped to transcripts derived from the gene models
using Bowtie.
}
\label{mapped-reads}
\end{figure}

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{chick_est_venn.pdf}
\end{center}
\caption{
    {\bf Splice junctions supported by ESTs or mRNAs.}
    Both Cufflinks and Gimme detect unique splice junctions
    supported by ESTs + mRNAs. A large number of splice junctions
    in ESTs + mRNAs are not found in either Gimme or Cufflinks
    because ESTs + mRNAs contain transcripts for other tissues.
}
\label{chick_est_venn}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{chick_venn.pdf}
\end{center}
\caption{
{\bf Splice sites in chicken Ensembl gene models detected by
Cufflinks and Gimme}
Cufflinks detects many annotated isoforms that are not detected
by the pipeline.  The figure also shows that both methods detect
a large number of unannotated splice junctions.}
\label{chick_venn}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{mouse_venn.pdf}
\end{center}
\caption{
{\bf Splice sites in mouse Ensembl gene models, Cufflinks and
Gimme pipeline.}
In contrast to chicken datasets, 85,160 (96.8\%) of splice
junctions found by Cufflinks are annotated in mouse ENSEMBL
models. While Gimme can detect the similar number of ENSEMBL
splice junctions, it detects many more non-ENSEMBL splice
junctions in mouse.}
\label{mus_venn}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{cuff_gimme_junctions_venn.pdf}
\end{center}
\caption{
{\bf Splice junctions found in merged models.} Merged models (B)
find more splice junctions in ENSEMBL and ESTs + mRNAs than
Cufflinks models only (A).}
\label{combined_venn}
\end{figure}

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{pipeline.pdf}
\end{center}
\caption{
{\bf Gene model construction pipeline.}
Transcripts are obtained from two assembly methods -- global and
local assembly. Transcripts with low complexity are removed and
poly(A) tails are trimmed by SeqClean. Transcripts are aligned to
the chicken genome using BLAT and alignments are then used by
Gimme to construct gene models. Gene models from Cufflinks and
other sources can also be incorporated to build the gene models.
}
\label{img:pipeline}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{local_assembly.pdf}
\end{center}
\caption{
{\bf Local assembly pipeline.}
Reads are first mapped to a chicken genome by Tophat.  Then reads
mapped to each chromosome are assembled by Velvet and Oases
separately.
}
\label{local_assembly}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{algorithm.pdf}
\end{center}
\caption{
{\bf Intron and exon graph.}
Introns sharing at least one exon are grouped together, then an
exon graph is constructed with exons from all introns as nodes.
Exons are excluded from the final gene model if they are smaller
than other exons sharing the same intron (exon 1a, 3a and 3b).
Edges of the exon graph (red arrow) are implied from the
connections of introns (blue arrow).  A transcript is constructed
from concatenation of all unique exons in the graph.
}
\label{algorithm}
\end{figure}
\end{document}

>>>>

Until recently, studies of alternative splicing have been limited to a
small number of genes and isoforms due to the high cost and low throughput
of sequencing expressed sequence tags (ESTs) and full-length cDNA
libraries.  RNA sequencing (RNA-Seq) using deep short-read sequencing
has been used successfully in many studies to gain unprecedented
insight into the complexity of transcriptomes (@cite).
It has been estimated that, in human, $92-94$\% of multiexon genes
undergo alternative splicing and different isoforms are expressed in
different tissues\cite{Wang:2008ea}.  This suggests that even in human
a large number of splice variants have not been explored.

@@

Despite the small size of sequencing reads, several studies have
detected novel splice junctions based on alignment of reads spanning
putative exon junctions.  To map reads across exon junctions,
reads are split into two parts and each part is mapped to the genome
independently (@cite).  A splice junction is then identified based on
alignments of each half of a read that falls between two exons at
exon-intron boundaries.  With this approach, Wang \emph{et al} have
identified a large number of splice junctions that are not annotated
from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.  These novel
splice sites include both canonical and non-canonical splice sites.
Approximately $46\%-75\%$ of canonical splice sites are supported by
ESTs.  Novel splice junctions have different levels of read coverage
suggesting that both high- and low-expressed isoforms are unannotated.
Using a similar approach, Pickrell {\em et al}~\cite{Pickrell:2010gt}
identified more than 150,000 novel canonical splice junctions in lymphoblastoid
cells.  The study also shows that the number of unannotated splice junctions
varies among cells from different human tissues, which suggests tissue-specific
expression of isoforms\cite{Pickrell:2010gt}.
% A majority of unannotated isoforms found in both studies are from alternative
% splicing of annotated exons. Only a small fraction of isoforms contain
% previously unknown exons.
% @CTB is this last point important?

@@

and Table~\ref{unique_sequences_matched_mouse}); this is probably due to the
incomplete nature of the current chicken genome assembly, which is lacking
approximately 5\% of its true gene content (@cite).  Interestingly, over a
hundred regions were present {\em only} in the local assemblies,
suggesting that the local assemblies might be recovering additional exons.
Significant numbers of unique regions from both global and local assemblies
showed homology to the mouse genome, indicating that at least some of these
unique sequences represented real sequence content.
Figure~\ref{global_vs_local} shows an example of different isoforms detected by
the two assembly methods.

@@

More importantly, we also mapped paired-end reads from technical replicates to
the same gene models, and found that more than 74\% of the paired-ends mapped
concordantly the gene models.  Most of the reads that did not map were either
highly erroneous or contained low-complexity artifactual sequence that probably
originated from sample processing and reverse transcription (@cite).  Thus the
merged gene models produced by Gimme represent the significant majority of the
assemblable data.

@@

Of 105,461 splice junctions from the gene models, 2,057 (2\%) junctions have no
spliced reads and only 4,626 (4.4\%) junctions have fewer than 4 spliced reads
(Fig.~\ref{cdf_splice}).  Note that 710 junctions are in chrUn\_random contigs,
which may have a great number of genome missassemblies. The number of junctions
outside random chromosomes with no reads is 1,347/99,986 (1.3\%).
% @CTB how do you get junctions with no spliced reads!?
% @LP our models have many 2,400 more junctions in random chromosomes than
% cufflinks and 710 of no reads junctions are from random chromosomes. It is
% likely that those junctions are from missassemblies.
% @LP Another reason is it has something to do with variation in read
% mapping between Tophat and bowtie.  Tophat maps spliced reads with less
% stringency because it has to allow gap alignment and it uses -v mode to map
% spliced reads. On the other hand, the default mode for bowtie is -n mode,
% which is the one I used.
% @LP I used a parameter that set max alignments to 100, only one alignment will
% be reported if there are more than 100 alignments. Maybe I should let
% bowtie report all alignments instead.
More than 95\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
real splice junctions.
% @CTB I'm slightly bugged by this statement.  Are we right?!
% @LP the main reason I use four reads as a cutoff because there are four
% samples and if at least one read from each sample map to the junction, it
% will likely be real.  Apparently, it is not always the case, but that is just
% my simple logic.

@@

Of the 105,461 splice junctions in our gene models, 83,560 (79.2\%) are
supported by ESTs or mRNAs from Genbank (Fig.~\ref{chick_est_venn}).  This is
especially surprising since our mRNAseq data is from spleen, and most of the
publicly available ESTs or mRNAs are from other tissues. 16,909 splice junctions
are found by either Cufflinks or our gene models, but not both, suggesting a variation in
sensitivity between two methods. Note that this cross-validation suggests that
the 12,188 novel splice junctions {\em not} seen in publicly available ESTs and
mRNAs are also likely to be real splice junctions from spleen.

@@

We next compared the Gimme gene models to those produced by Cufflinks,
another reference-based approach to building gene models from mRNAseq
data\cite{Trapnell:2010kd}.  We also compared the results from both
methods to the ENSEMBL gene annotations, which are produced by a
pipeline that incorporates de novo gene prediction and homology-based
approaches as well as expression data (@cite).

@@

When we apply Gimme and Cufflinks to a publicly available mouse mRNAseq data
set, Gimme and Cufflinks recover approximately the same number of splice
junctions already known from ENSEMBL (Figure ~\ref{mus_venn}).  However, Gimme
recovers a substantial number of additional splice variants beyond Cufflinks and
ENSEMBL both.

@@

As shown above, Cufflinks and the Gimme method detect a number of distinct but
equally valid splice junctions, which suggests that we could obtain greater
sensitivity to exon-exon junctions in our gene models by merging both sets of
predictions.  We therefore used Gimme to incorporate the Cufflinks models to
global and local assembly, Table.~\ref{genes_transcripts}.  This resulted in a
decreased number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{}). The merged gene models recover 49.3\% and 56.2\% of
splice junctions from ESTs and ENSEMBL respectively
(Fig.~\ref{combined_venn}), which is about 10\% greater than that from
corresponding unmerged gene models.

@@

% @LP Maybe I should add bitscores from Cufflinks so that we can compare with
% those from Gimme. If the results are similar, we can say that extending
% Cufflinks with assembly does not affect gene structures in a bad way.
To validate our predicted isoforms, we extracted putative coding sequences from
our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan successfully
translated 28,772 of 34,800 (82.6\%) of our isoforms to protein sequences with
50 or more amino acids.  We then searched for homologous sequences in mouse
ENSEMBL, and found that 22,991 (79.9\%) of our isoforms from 12,945 distinct
genes match mouse proteins at a bit-score/length $\ge1.0$
(Fig.~\ref{bitscore}).  A bit-score/length ratio greater than 1 indicates that
approximately more than half of the protein sequence matches mouse proteins
well. The results suggest that merging gene models from Cufflinks and {\em de
novo} aseembly did not disrupt the structure of protein sequences.

% @CTB for discussion: what's surprising about all of this? so many
% splice variants, etc?  Note we need to go over pickrell paper carefully.
% @LP I discuss about it below. Hope it helps.

@@

\subsection*{{\em De novo} assembly should be used to extend gene models from
Cufflinks}
For organisms with a reference genome and ENSEMBL annotation, Cufflinks with
ENSEMBL gene models as a reference guide seems to be a preferable method for
building gene models. However, we have shown that ENSEMBL models in chicken are
missing a substantial number of splice variants found in ESTs and mRNAs that
are also not recovered by Cufflinks. Moreover, results from mouse data suggest
that Tophat's default parameters may have been tailored to maximize the finding
of isoforms in ENSEMBL models (Fig.~\ref{mus_venn}), which could lead to
restricted detection of novel splice variants.
% 85,512 (97.2\%) of splice
% junctions found by Cufflinks are also canonical compared to 92.6\% in mouse
% ENSEMBL gene models. Assuming that mouse data were not selected for splice
% variants with canonical splice junctions, the percentage of canonical splice
% sites found by Cufflinks is rather high. 
Therefore, gene models built from RNA-Seq reads using Cufflinks + ENSEMBL gene
models alone may not be suitable for genes and isoforms analyses that include
finding novel genes and splice variants.  {\em De novo} assembly is not tied to
any specific gene model parameters and does not only detect canonical splice
sites, thus we recommend that it is used to extend gene models built from
Cufflinks + ENSEMBL gene models, especially in organisms with relatively
incomplete ENSEMBL gene models.  Besides {\em de novo} assembly,
annotation from other sources can also be incorporated to extend Cufflinks models
using our pipeline.

@@

\subsection*{Mapping reads to The Genome and Gene Models}

@@

% @LP check the version of tophat
Reads from each dataset were first assembled separately by global assembly
without using a reference genome.  In contrast, reads from each dataset were
first mapped to the chicken genome using Tophat2.  Then only reads mapped to
the genome were assembled by chromosome in the local assembly
(Fig.~\ref{local_assembly}).  Global and local assembly was performed using
Velvet version 1.2.03\cite{Zerbino:2008vu} with default parameters except for
hash length (k-mer).  A range of k-mer length from 21-31 was used to assemble
reads from chicken data and k-mer length 27 was used to assemble reads from
mouse data.  Lastly, transcripts from both methods were assembled by Oases
version 0.2.06\cite{Schulz:2012je}.

@@

Figure~\ref{overall_pipeline} depicts an overall gene model construction
pipeline.  Transcripts of all datasets from local and global assembly were
mapped to the chicken genome using BLAT\cite{Kent:2002tv}. Alignments and gaps
from BLAT outputs are considered exons and introns respectively.  Optionally,
data from other sources (ESTs, RefGenes, etc.) can be incorporated with
transcripts from the assembly to improve gene models.  All transcripts are then
assembled using Gimme, a program that assembles transcripts based on their
alignments to the reference genome.  An algorithm for assembling transcripts is
described below.  A maximum set of transcripts obtained from Gimme are then
reduced to only a minimum set of transcripts that contain all splice junctions
and untranslated regions (UTRs).

@@

The algorithm first builds an intron graph using introns as nodes.  Each intron
contains exons whose splice sites perfectly match intron boundaries. At the
$5\prime$ and the $3\prime$ ends, some exons may be smaller than others due to
incomplete assembly (supp. Fig.??). Only the largest exons are kept to ensure
that the transcript has the most complete exons (Fig.~\ref{algorithm}).
Transcripts were then grouped into the same gene if they have at least one
intron or exon in common.  Then, a splice graph composed of exons is created
and structures of isoforms are derived from traversing paths in the splice
graph.  Gimme is open-source and available at
\texttt{https://github.com/ged-lab/gimme}.

@@

\section*{Figure Legends}
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{overall_pipeline.pdf}
\end{center}
\caption{
{\bf Gene model construction pipeline.} Transcripts are obtained from two
assembly methods -- global and local assembly.  Transcripts are aligned to the
chicken genome by BLAT\@. Gimme then constructs gene models based on alignments
of transcripts.  }
\label{overall_pipeline}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{local_assembly}
\end{center}
\caption{
{\bf Local Assembly Pipeline.}
Reads are first mapped to a chicken genome.  Then only mapped reads are
assembled by Velvet and Oases.
}
\label{local_assembly}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{algorithm.pdf}
\end{center}
\caption{
{\bf Intron and exon graphs.} Each intron connects to exons whose splice
junctions match it boundary.  Some exons are excluded from the final gene model
if they are incomplete (exon 3a,b).  Introns sharing at least one exon are
grouped together.  Then an exon graph is made using exons as nodes.
}
\label{algorithm}
\end{figure}

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{mapped-reads.pdf}
\end{center}
\caption{
{\bf Cumulative counts of splice junctions with spliced reads from both single-
and paired-end data from Cufflinks and Gimme models.}
}
\label{mapped-reads}
\end{figure}

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{chick_est_venn.pdf}
\end{center}
\caption{
    {\bf Splice junctions supported by ESTs or mRNAs.}
}
\label{chick_est_venn}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{chick_venn.pdf}
\end{center}
\caption{
{\bf Splice sites in chicken Ensembl gene models detected by Cufflinks and the
{\em de novo} assembly pipeline.} Cufflinks detects many annotated isoforms
that are not detected by the pipeline.  The figure also shows that both methods
detect a large number of unannotated splice junctions.}
\label{chick_venn}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{mouse_venn.pdf}
\end{center}
\caption{
{\bf Splice sites in mouse Ensembl gene models, Cufflinks and the \emph{de
novo} assembly pipeline.} In contrast to chicken datasets, 85,160 (96.8\%) of
splice junctions found by Cufflinks are annotated in mouse ENSEMBL models. The
higher percentage of ENSEMBL splice junctions found by Cufflinks may be a
result of more complete ENSEMBL gene models. While Gimme can detect the similar
number of ENSEMBL splice junctions, it detects many more non-ENSEMBL splice
junctions in mouse.}
\label{mus_venn}
\end{figure}

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{cuff_gimme_junctions_venn.pdf}
\end{center}
\caption{
    {\bf Splice junctions found in merged models.} Merged models (B) find more
    splice junctions in ENSEMBL and ESTs + mRNAs than Cufflinks models only (A).
}
\label{combined_venn}
\end{figure}

@@

\end{document}
